Visualisation of Brain Statistics with R-package \texttt{ggseg}

### 2019-11-28 test.Rmd
revised based on didacs comments

---
title: Visualisation of Brain Statistics with R-package `ggseg`
preprint: true
author: 
  - name: Athanasia M. Mowinckel
    affiliation: 1
    corresponding: true
    email: a.m.mowinckel@psykologi.uio.no
  - name: Didac Vidal Piñeiro
    affiliation: 1
affiliation:
  - code: 1
    address: Center for Lifespan Changes in Brain and Cognition, Univeristy of Oslo, PO. box 1094 Blindern, 0317 Oslo, Norway
abstract: >
  The abstract of the article.
  It can also be on *multiple* lines.
bibliography: references.bib
output:
  bookdown::pdf_book:
    base_format: rticles::peerj_article # for using bookdown features like \@ref()
    citation_package: natbib
    fig_caption: true
    includes:
      in_header: preamble.tex
  rticles::peerj_article: preprint
---
```{r setup, include=F}
knitr::opts_chunk$set(eval=TRUE, echo=TRUE)
options(tinytex.verbose = TRUE)

@@

MRI data provides good spatial resolution and thus an optimal representation has to respect spatial relationships across regions.
Results from brain atlas analyses are most meaningfully visualized when projected onto a representation of the brain, rather than on other types of graphs (e.g. bar charts).
The projection of data into brain representations provides clear points of reference - especially when the reader is unfamiliar with the atlas - eases readability, guides interpretation, and informs on the spatial patterns of the data.
Adopting the grammar of graphics implemented in ggplot2 (@ggplot), one can plot neuroimaging data directly in R with several tools such as ggBrain (@ggBrain) and ggneuro (@ggneuro; see neuroconductor -@neuroconductor for curated neuroimaging packages for R).
Yet these tools display whole-brain image files and are not well-suited for representing brain atlas data.

@@

The ggseg package presents both compiled data sets, tailored scripts that allow brain data integration and plotting and, other minor features such as custom color palettes.
The data featured in the ggseg package are derived from two well-known parcellations: the Desikan-Killany cortical atlas (DKT; @dkt), which covers the cortical surface of the brain, and the Automatic Segmentation of Subcortical Structures (aseg; @aseg), which covers the subcortical structures.
Both atlases are implemented in several neuroimaging softwares, such as FreeSurfer (@fischl_99, @dale_99, @Fischl2000), and are commonly used in relation to developmental changes, disease biomarkers, genomic data, and cognition (@amlien_elaboration_2019, @WALHOVD20051261, @Pizzagalli).
The ggsegExtra package includes a library of precompiled atlases (currently including `r dd %>% filter(Package == "ggsegExtra") %>% nrow()` additional atlases) and it is updated continuously.
See Table \@ref(tab:atlasTab) for a summary of the available atlases (as for May 2019).
We encourage users to contribute to the ggsegExtra brain atlas repository by including additional brain atlases.
In SI text and [github wiki](https://github.com/LCBC-UiO/ggseg/wiki), we offer a pipeline to create and supply atlases for 2D plotting.
See SI text and SI scripts for instruction to include atlases for 3D plotting.

```{r "atlasTab", echo=F, warning=FALSE, message=FALSE}
dd %>%
  kable(format="latex", escape = FALSE, booktabs = TRUE,
               caption = "Table of currently available atlases in the ggseg and the ggsegExtra R-packages. Polygon and mesh refer to 2D and 3D brain atlase representations.") %>%
  collapse_rows(1)
```

# Tutorial
This tutorial will introduce the `ggseg` and `ggsegExtra` packages and familiarize the reader with the main functions and the general use of the packages.
The tutorial will focus on the two main functions: `ggseg()` for plotting 2D polygons and `ggseg3d()` for plotting 3D brains based on tri-surface mesh plots.

## Plotting polygon data (ggplot2)

`ggseg` is the main function for plotting 2D data.
By default, the function automatically plots the DKT atlas (see Figure \@ref(fig:init)).
The `ggseg` function is a wrapper for `geom_polygon` from `ggplot2`, and it can be built upon and combined like any ggplot object.
The image plot consists of a simple brain representation containing no extra information.
Hence, `ggseg` plots can be easily complemented with any of the available ggplot features and options.
We recommend users to get familiarized with ggplot2 grammar (@ggplot).

```{r "init", fig.width=3, message=FALSE, fig.cap="By default `ggseg` will plot the dkt atlas in shaded polygons."}
library(ggseg)
library(tidyverse)
ggseg()
```

As `ggseg` is implemented as a fully-functional part of ggplot, one can combine the function together with any other ggplot function such as themes or scales.
Also, one has several options for plotting the main brain representations.
These options are atlas-specific.
For cortical atlases, such as the 'dkt', one can stack the hemispheres, view only the medial or lateral side, choose either one or both hemispheres, or any combination of hemisphere and view.
See Figure \@ref(fig:collection)) for an example.
For subcortical atlases, such as the 'aseg', the options are more limited but one can often choose between axial, sagital, and coronal views.

```{r collection, fig.width=5, fig.cap="ggseg plots can be used with any ggplot feasture such as standard scales and themes. For cortical atlases, one can supply special ggseg options to determine hemisphere or view, and control how the plots looks like and is organised. **A:** dkt atlas, stacked with dark theme ; **B:** dkt with medial view only; **C:** dkt atlas with only left medial display; **D:** dkt atlas, stacked, with classic theme; **E:** dkt atlas with left hemisphere only; **F:** complete aseg atlas"}
# dkt dark theme
p1 <- ggseg(position = "stacked") +
  theme_dark() +
  labs(title=" ")

# dkt classic theme
p2 <- ggseg(position = "stacked") +
  theme_classic() +
  labs(title = " ")

# dkt medial view
med <- ggseg(view = "medial") +
  labs(title = " ")

# dkt left hemisphere
left <- ggseg(hemisphere = "left") +
  labs(title = " ")

# aseg default theme
p3 <- ggseg(atlas=aseg) +
  labs(title = " ")

# dkt left medial alone
combo <- ggseg(view = "medial",
      hemisphere = "left") +
  labs(title=" ")

# Combine plots
cowplot::plot_grid(p1, med, combo,  p2, left, p3,
                   labels = c("A: dkt - dark", "B: dkt - medial",
                              "C: dkt - combo", "D: dkt - classic",
                              "E: dkt - left", "F: aseg"),
                   hjust = -.05)

```

### Using own data with fill and colour
`ggseg` accepts any argument you can supply to `geom_polygon` and therefore is easy to work with for those familiar with ggplot functionality.
Standard arguments like `fill` that floods the segments with a colour, or `colour` that colours the edges around the segments are typical arguments to provide to the function either as a single value or within the ggplot mapping option `aes`.
To use color palettes corresponding to those used in the original neuroimaging softwares one can use atlas-specific 'brain' palette scales.
See an example in Figure \@ref(fig:fill).

```{r fill, fig.width=5, fig.height=8, fig.cap="Supplying 'area' to the fill option in ggseg, will use the column 'area' from the accompanying dataset to create a discrete colour palette over the segments in the atlas. The 'dkt' atlas default palette corresponds to the FreeSurferColorLut scheme."}
ggseg(mapping=aes(fill = area), colour="black") +
  scale_fill_brain("dkt") +
  theme(legend.justification=c(1,0),
        legend.position="bottom",
        legend.text = element_text(size = 5)) +
  guides(fill = guide_legend(ncol = 3))
```

Most users will use ggseg to display - using a color scale - some descriptive or inferential statistics, such as mean thickness or brain-cognition relationships across the different brain regions.
Yet, before projecting the statistics onto the segments, we need to explore the structure of the atlas data sets. The atlas dataset structure will help us understand what incoming statistical data needs to look like.
Note that each atlas corresponds to a unique data set. All data sets have a similar structure and contain key information regarding the atlas, the region names, and the coordinates for the segment polygons.

```{r, fig.cap="All ggseg polygon atlases are required to have at minimum 6 columns with the names: 'long', 'lat', 'id', 'hemi', 'area', 'side'. These columns are necessary for the ggseg function to work. The additional columns in the dkt were added for various practical reasons.", size='small'}
dkt
```

In any atlas, the column 'label' is particularly useful for combining the data of interest with the `ggseg`-polygons.
The column 'label' contains the label (region) names as in the original neuroimaging software.
For example, the DKT atlas label column matches the region names from Freesurfer statistics table outputs.
Yet the data in `ggseg` is in a long format - that is each region has its own row - and any data of interest needs to be in this same format.
Often data sets are organized in wide format, in which subjects are represented by rows and each different data variable is represented in a separate column, and thus need to be rearranged in order to work with ggseg.
See below an example of wide-to-long conversion.

```{r }
freesurfer_stats <- data.frame(
  id = c(10:12),
  lh_superiortemporal = c(3.32, 4.1, 3.5),
  lh_precentral = c(2.3, 2.5, 2.1),
  lh_rostralmiddlefrontal = c(3.3, 3.2, 3.1)
  )
freesurfer_stats

freesurfer_long <- freesurfer_stats %>%
  gather(label, thickness, -id)
freesurfer_long
```

Data in long format can then be used directly with the `ggseg` function, as the `label` column corresponds in name and content with the `label` column in the atlas data of dkt.
The data **must** include a column that has the same name and at least _some_ data matching the values in the corresponding column in the atlas data.
In the next example we create some data with 4 rows, and an 'area' and 'p' column, representing the results of a hypothetical analysis.
The `ggseg` function will recognise the matching column 'area', and merge the supplied data into the atlas using `dplyr` joins.
We use the `p` column as the column flooding the segment with colour.

```{r datasupp, fig.width=5}
someData = data.frame(
  area = c("transverse temporal", "insula",
           "pre central","superior parietal"),
  p = sample(seq(0,.5,.001), 4),
  stringsAsFactors = FALSE)

p <- ggseg(.data=someData, mapping=aes(fill=p))
p
```

The appearance of this plot can then be modified similarly to any other ggplot graph using functions such as scales, labs, themes, etc.

```{r datasupp2, fig.width=5}
p +
  theme_void() +
  scale_fill_gradient(low="firebrick",high="goldenrod") +
  labs(title="A nice plot title", fill="p-value")
```

If the results are only in one hemisphere, but you still want to plot both of them, make sure your data.fame includes the column `hemi` with either 'right' or 'left' for this to happen. In this case, data will be merged into the atlas both by 'area' and by 'hemi'.

```{r datasuppX, fig.width=5}
someData$hemi = "right"

ggseg(.data=someData, colour="white",mapping=aes(fill=p)) +
  theme_void() +
  scale_fill_gradient(low="firebrick",high="goldenrod") +
  labs(title="Another nice plot title")
```

### Creating subplots
There is often the need to plot a statistic of interest in different groups (e.g. thickness or brain - cognition relationships in young or older adults).
This may be obtained also with ggseg, using ggplot's `facet_wrap` or `facet_grid`, using two guiding rules:
**1)** as before, data needs to be in long format (group data should appear in seperate rows, not in separate columns).
**2)** The data needs to be grouped using `dplyr`'s `group_by` function _before_ providing the data to the `ggseg` function.
The ggseg function will detect grouped data, and adapt it to `facet`'s requirements.

```{r datasupp3, fig.width=5}
someData = data.frame(
  area = rep(c("transverse temporal", "insula",
               "pre central","superior parietal"),2),
  p = sample(seq(0,.5,.001), 8),
  AgeG = c(rep("Young",4), rep("Old",4)),
  stringsAsFactors = FALSE) %>%
  group_by(AgeG)

ggseg(.data=someData, colour="white", mapping=aes(fill=p)) +
  facet_wrap(~AgeG, ncol=1) +
  theme(legend.position = "bottom")
```

As before, one can apply the different ggplot and facet options, such as `scales` to modify the plot's appearance.

```{r datasupp5, fig.width=5}
ggseg(.data = someData, atlas=dkt,
      colour="white", position="stacked",
      mapping=aes(fill=p)) +
  facet_wrap(~AgeG, ncol=2) +
  theme_dark() +
  theme(legend.position = "bottom",
        axis.text = element_blank(),
        axis.title = element_blank()
        ) +
  scale_fill_gradient2(high="goldenrod",
                       low="firebrick",
                       na.value="transparent")
```

All the concepts described above also work with the 'aseg' atlas for subcortical structures, with the exception of 'hemisphere' and 'view' arguments that are superfluous in subcortical atlases such as aseg.
Again, inspecting the atlas data directly will be of aid when preparing data to use with the atlas. 
Note that both the equivalent structures between 'aseg' and the 'dkt' atlases.

```{r , size = 'small', out.width='90%'}
aseg
```

```{r atlases, fig.width=5, fig.cap="The first 10 rows of the `aseg` dataset, that has some specific differences from the dkt, like missing the option to not view both hemispheres, as the plot would be hard to understand that way."}
ggseg(atlas="aseg", mapping=aes(fill=area))
```

## Plotting 3D mesh data

Representing brains as 2D polygons is a good solution for fast, efficient, and flexible plotting, and can be easily combined with interactive apps such as Shiny (@shiny).
Yet, brains are intrinsically 3-dimensional and it can be challenging to recognize the location of a region as a flattened image.
This problem is exacerbated in atlases that represent subcortical features – after all, cortical surfaces are 2-dimensional – such as grey matter structures or white matter tracts.
Hence, here we also provide the `ggseg3d` function to plot, view, and print 3D-atlases in R.
`ggseg3d` is based on tri-surface mesh plots using plotly (@plotly).
The data structure is somewhat more complex than the ggplot polygons, and includes additional options for brain inflation, glass brains, camera locations, etc.
As `ggseg3d` is based on plotly, the resulting brain atlases are interactive, which guides interpretation, and is useful for public dissemination.
We recommend users to familiarize themselves with plotly (@plotly) when using this function.

Out-of-the-box, `ggseg3d()` plots the `dkt_3d` atlas in 'LCBC' surface.
The 'LCBC' surface consists on a semi-inflated white matter surface based on the _fsaverage_ template subject.
See *SI SCRIPT* for reproducibility.
All `[...]_3d` atlases have a built in `colour` column for default colour plotting of the regions, based on the color scheme used in the neuroimaging software.

```{r ggseg3d-1-out, message=FALSE, echo=F, eval=TRUE, out.width='30%',fig.show='hold',fig.ncol = 3 , fig.cap="The three surface options provided in ggseg 3d atlases. From left: the 'white' surface is the white matter surface, 'LCBC' surface is the semi-inflated white matter surface (10 iterations), and the 'inflated' surface is a fully inflated white matter cortical surface."}
p <- ggseg3d(surface = "white")  %>%
  layout(title="\nggseg3d(surface='white')",
         font =  list(size = 30))
orca(p, file = "png/ggseg3d_white.png", format = "png")

p <- ggseg3d()  %>%
  layout(title="\nggseg3d(surface='LCBC')",
         font =  list(size = 30))
orca(p, file = "png/ggseg3d.png", format = "png")

p <- ggseg3d(surface = "inflated") %>%
  layout(title="\nggseg3d(surface='inflated')",
         font =  list(size = 30))
orca(p, file = "png/ggseg3d_inflated.png", format = "png")

include_graphics(c("png/ggseg3d_white.png","png/ggseg3d.png", "png/ggseg3d_inflated.png"))
```

The 3D-atlas data is stored in nested tibbles.
Each cortical atlas has data sets for three different surfaces (see Figure \@ref(fig:ggseg3d_1_out)) and the two hemispheres.
Only one surface is available for subcortical atlases as inflation procedures do not yield better representations.
The 'ggseg_3d' column includes all necessary information to `ggseg3d()` to create a 3D mesh-plot.
The additional 3D-atlases in ggsegExtra have the same data structure.

```{r, out.width='90%', size='small'}
dkt_3d
```

### External data supply

Similarly as in the 2D-atlas, the user will use `ggseg3d` to display through a color scale some descriptive or inferential statistics.
If the data is not already in the correct long format, or uses similar naming as the atlas, one needs to grab the data for a specific surface (and hemisphere, if desired) and then `unnest(ggseg_3d)` it to see what the atlas expects.

```{r out.width='40%', linewidth=70}
dkt_3d %>%
  filter(surf == "inflated" & hemi == "right") %>%
  unnest(ggseg_3d) %>%
  select(-lobe, -acronym)
```

Note the `mesh` column, which contains lists.
Each list corresponds to a region and contains 6 vectors required to create the mesh of the tri-surface plot.
It should also be noted that the 'label', 'annot' and 'area' columns could provide matching values for your own data.
Similarly to the `ggseg`-function, the 'label' column should match the region names used in the original neuroimaging software while 'area' and 'annot' provide alternative/secondary names.
It is thus important to match your regional identifiers with those used in the atlas.
A warning will be issued in case of any mismatch.
The column you want to use for colour, needs to be supplied to the `colour` option, and you'll likely want to supply it to the `text` option, as this will add another line to the plotly hover information.

```{r, eval=FALSE, out.width='40%'}
someData = dkt_3d %>%
  filter(surf == "inflated" &
         hemi == "right") %>%
  unnest(ggseg_3d) %>%
  select(area) %>%
  na.omit() %>%
  mutate(p = sample(seq(0,.5, length.out = 100 ), nrow(.)) %>%
           round(2))

  ggseg3d(.data = someData, atlas = dkt_3d, colour = "p", text = "p")
```

```{r, out.width='40%', echo=FALSE}
someData = dkt_3d %>%
  filter(surf == "inflated" & hemi == "right") %>%
  unnest(ggseg_3d) %>%
  select(area) %>%
  na.omit() %>%
  mutate(p = sample(seq(0,.5, length.out = 100 ), nrow(.)) %>%
           round(2))

p <- ggseg3d(.data = someData, atlas = dkt_3d,
             colour = "p", text = "p")
include_plotly(p, "png/ggseg3d_data.png")
```

### Colours

You can provide custom colour palettes either in hex or R-names.
Colours will be evenly spaced when creating the colour-scale, and you may provide as many colours as you wish.

```{r,fig.width=5, eval=FALSE, out.width='45%'}
  ggseg3d(.data = someData, atlas = dkt_3d,
          colour = "p", text = "p",
          palette = c("#ff0000", "#00ff00", "#0000ff"))
```
```{r, out.width='40%', echo=F, eval=T}
p <- ggseg3d(.data = someData, atlas = dkt_3d,
          colour = "p", text = "p",
          palette = c("#ff0000", "#00ff00", "#0000ff"))
include_plotly(p, "png/ggseg3d_colour.png")
```

If you want to alter the colour of `NA` regions, supply `na.colour`, either as HEX colour or colour name.
This option only takes a single colour.

```{r, out.width='40%', eval=FALSE,  out.width='45%'}
  ggseg3d(.data = someData, atlas = dkt_3d,
          colour = "p", text = "p",
          na.colour = "black", camera="medial")
```
```{r, out.width='40%', echo=F, eval=T}
p <-   ggseg3d(.data = someData, atlas = dkt_3d,
          colour = "p", text = "p",
          na.colour = "black", camera="medial")
include_plotly(p, "png/ggseg3d_nacolour.png")
```

Subcortical atlases include cortical surfaces and other landmark structures for visualization purposes only.
One can control the opacity of the these `NA` structures, to improve visualization.
Additionally, one can add the `glassbrain` option, which provides a frame of reference for the subcortical structures.
Glassbrains can be controlled with three different arguments: opacity (0-1), hemisphere (left, right), and colour.

```{r, out.width='40%', warning=F, message=FALSE, eval=FALSE}
somData_aseg = aseg_3d %>%
  unnest() %>%
  select(label) %>%
  filter(!grepl("Ventricle|Putamen", label)) %>%
  mutate(p = seq(1, nrow(.)))

  ggseg3d(.data = somData_aseg, atlas = aseg_3d,
          colour = "p", text = "p",
          na.alpha= .5, glassbrain = .5,
          glassbrain_hemisphere = "left")
```
```{r, out.width='40%', echo=F, eval=T}
somData_aseg = aseg_3d %>%
  unnest() %>%
  select(label) %>%
  filter(!grepl("Ventricle|Putamen", label)) %>%
  mutate(p = seq(1, nrow(.)))

p <- ggseg3d(.data = somData_aseg, atlas = aseg_3d,
          colour = "p", text = "p",
          na.alpha= .5, glassbrain = .5,
          glassbrain_hemisphere = "left")
include_plotly(p, "png/ggseg3d_glass.png")
```

`ggsed3d()` is based on plotly and thus additional plotly functionalities can be used to modify and improve the 3D atlas representations.
In addition to Carson Sievert's book on plotly in R (-@plotly), we recommend resources for modifying axes in 3D plots (@plotly-ax), the basic introduction to tri-surface plots (@plotly-tri), and this tutorial on tri-surface plots with plotly in R (@plotly-trisurf).
Finally, we recommend [orca](https://github.com/plotly/orca#installation) command line tool to save ggseg3d atlas snapshots.

# Discussion

The main aim of the ggseg – and ggsegExtra - package is to ease and streamline visualization of brain atlas data in R, by gathering a collection of atlases from several scientific sources and providing plotting functions adapted to these atlases.
In this tutorial, we introduce the package to the readers by presenting some use examples and highlighting the main functions and options that are available.
As a visualization tool, this package adds up to manifold functionalities such as ggBrain (@ggBrain) and ggneuro (@ggneuro) in R, and software-specific image viewers such as FSLeyes (@fsleyes) and Freeview (@dale_99).
In this regard, we do not aim to compete with software-specific visualizations or advocate for the superiority of ggseg as a visualization tool.
After all, flattened 2D polygons do not rely on a meaningful brain coordinate system and the units of information in 3D meshes are limited to the number of parcellations.
On the contrary, we believe the ggseg niche among visualization tools resides in its simplicity and its ability to be combined with statistical analysis pipelines.
The possibility to serve as an interactive tool for dissemination and reproducibility when combined with other technologies, such as Binder (@binder) or Shiny (@shiny), is an added benefit.

The ggseg and ggsegExtra packages contain three main features:
**1)** a collection of 2D-polygon and 3D mesh brain parcellation atlases. The atlas data are necessary to create the coordinate system for the brain atlas plotting, but are made in such a way to make it easy to use for ones own purposes.
**2)** `ggseg()` and `ggseg3d()` functions for visualization.
`ggseg()` is a wrapper function for `geom_polygon` from `ggplot2` and it can be built upon like any ggplot object.
`ggseg3d()` is a plotly wrapper function for tri-surface mesh plots which prints 3D atlases.
Both functions are flexible and well-adapted to its environment as can be combined with any additional argument from ggplot and plotly, respectively.
**3)** Complimentary features – e.g. color scales - and functions such as `as_ggseg_atlas()` and `as_ggseg3d_atlas()` to convert data in the correct atlas format. These function should provide users with the possibility of adapting their plots to their wishes, and also to make it possible to create and contribute to the atlas repository.

The foundations of the ggseg-package trace back to the necessity of visualizing and exploring the lifespan trajectories of cortical thickness across different brain regions (see supporting information in Vidal-Piñeiro et al. (-@vidal_2019)) .
That is, ggseg appears with the need to inspect and display brain information over time - that is, including a spatial diemension and a time-varying factor - overcoming the constrains of printed journals and classical 2D plots (e.g. bar plots).
The current state of science requires researches to share the results of studies in both high detail and in an intuitive manner, as it permits communication to wide audiences and facilitates reproducibility.
Hence, we believe this tool conforms to the essence of open science and invite users to improve the code, provide examples, or tutorials, and contribute to the atlas collection according to their own interest and needs via the public [ggseg GitHub repository](https://github.com/LCBC-UiO/ggseg) and [ggsegExtra GitHub repository](https://github.com/LCBC-UiO/ggsegExtra).
See X and Y for detailed guidelines on how to create new atlases.
Finally - while the ggseg-package is circumscribed to brain parcellations - we believe that the structure and functions of the package can be easily applied to any scientific field that benefits from data being displayed across the spatial dimension.
We encourage readers to borrow the package functionalities and adapt it to their respective fields and structures of interest, such as has already been done with the gganatogram-package (@gganatogram).

# Conclusion
Visualization is a fundamental aspect of neuroimaging to explore and understand data, guide interpretation and, communicate with colleagues and general audience.
In this tutorial, we have introduced the ggseg-package, a tool for visualizing brain statistics through brain parcellation atlases in R.
This visualization tool easily combines with interactive routines as well as with diverse statistical analysis pipelines. We hope this tool and tutorial proves useful to neuroscientists and inspires others to apply the functions in a wide variety of fields and structures.

# Author Contributions
Didac Vidal-Piñeiro generated the idea for the tool, and the initial scripts that made the plots.
He has also been responsible for converting images from neuroimaging software into the type of data necessary for plotting polygons and mesh plots in R.
Athanasia M. Mowinckel adapted the initial scripts and made the functions into package format and has continued developing the functions with the aim of increasing user-friendliness.
She is also responsible for conceiving and adding the mesh-plot functionality through plotly, and developing the pipeline for making that possible.
A. M. Mowinckel wrote the first draft of the paper, and both have since critically edited it.

# Conflicts of Interest
The authors declare that there were no conflicts of interest with respect to the authorship or the publication of this article.

# Acknowledgements
We wish to thank John Muschelli for his initial improvements in the package code and preparing it for incorporation in neuroconductor (-@neuroconductor), and Richard Beare for submitting the first community created atlas, and providing helpful comments to the base code.
A. M. Winkler, who originally introduced us to the procedure of converting 3D brain images from Freesurfer into `.ply` files, also needs to be credited with this, we would not have managed to create the mesh-plots without his scripts.
We also thank Anders M. Fjell and Kristine B. Walhovd for allowing us to pursue the creation of this tool and encouraging our progress.
We are indebted to those users that actively contributed to the package and all the scientists and engineers that worked on the framework on which the package is sustained (e.g. R-project, neuroimaging software).

# Funding
This work is funded by EU Horizon 2020 Grant 'Healthy minds 0-100 years: Optimizing the use of European brain imaging cohorts (Lifebrain)', with grant agreement number 732592.

# Prior versions
Some of the content presented here also appears in the package vignette of `ggseg`, which may be accessed through R or in the package website (@ggseg).
Athanasia Monika Mowinckel also has several tutorials on her blog regarding ggseg creation and functionality (@ggsegAnim, @ggsegIntro).

# References

>>>>

---
  title: Visualisation of Brain Statistics with R-package `ggseg`
preprint: true
author: 
  - name: Athanasia M. Mowinckel
affiliation: 1
corresponding: true
email: a.m.mowinckel@psykologi.uio.no
- name: Didac Vidal Piñeiro
affiliation: 1
affiliation:
  - code: 1
address: Center for Lifespan Changes in Brain and Cognition, Univeristy of Oslo, PO. box 1094 Blindern, 0317 Oslo, Norway
abstract: >
  There is an increasing emphasis on visualising results from statistical analysis in more intuitive ways. Neuroimaging data already have, as a consequence of the source of the data, particular shapes and as such visualisation of these data should reflect these same dimensional properties. The ggseg package makes it possible to visualise pre-defined brain segmentations of local/structural or connectivity features in both 2D polygons or 3D mesh. Presenting both relative position and size of segments in addition to statistical summary information aides interpretation and dissemination of results. In this tutorial, we present the data and functions in the ggseg R-package for brain atlas visualisation. The tutorial highlights the main functions to create brain segmentation plots in R, and also introduces the accompanying package data that makes these visualisations possible. 
bibliography: references.bib
output:
  bookdown::pdf_book:
  base_format: rticles::peerj_article # for using bookdown features like \@ref()
citation_package: natbib
fig_caption: true
includes:
  in_header: preamble.tex
rticles::peerj_article: preprint
---
  ```{r setup, include=F}
knitr::opts_chunk$set(eval=TRUE, echo=TRUE)
options(tinytex.verbose = TRUE)

@@

MRI data provides good spatial resolution and thus an optimal representation has to respect spatial relationships across regions.
Results from brain atlas analyses are most meaningfully visualized when projected onto a representation of the brain, rather than on other types of graphs (e.g. bar charts) alone.
The projection of data onto brain representations provides clear points of reference - especially when the reader is unfamiliar with the atlas - eases readability, guides interpretation, and conveys the spatial patterns of the data.
Adopting the grammar of graphics implemented in ggplot2 (@ggplot), one can plot neuroimaging data directly in R with several tools such as ggBrain (@ggBrain) and ggneuro (@ggneuro; see neuroconductor -@neuroconductor for curated neuroimaging packages for R).
Yet these tools display whole-brain image files and are not well-suited for representing brain atlas data.

@@

The ggseg package presents both compiled data sets, tailored scripts that allow brain data integration and plotting, and other minor features such as custom color palettes.
The data featured in the ggseg package are derived from two well-known parcellations: the Desikan-Killany cortical atlas (DKT; @dkt), which covers the cortical surface of the brain, and the Automatic Segmentation of Subcortical Structures (aseg; @aseg), which covers the subcortical structures.
Both atlases are implemented in several neuroimaging softwares, such as FreeSurfer (@fischl_99, @dale_99, @Fischl2000), and are commonly used in relation to developmental changes, disease biomarkers, genomic data, and cognition (@amlien_elaboration_2019, @WALHOVD20051261, @Pizzagalli).
The ggsegExtra package includes a library of precompiled atlases (currently including `r dd %>% filter(Package == "ggsegExtra") %>% nrow()` additional atlases) and it is updated continuously.
See Table \@ref(tab:atlasTab) for a summary of the available atlases (as for May 2019).
We encourage users to contribute to the ggsegExtra brain atlas repository by including additional brain atlases.
In SI text and [github wiki](https://github.com/LCBC-UiO/ggseg/wiki), we offer a pipeline to create and supply atlases for 2D plotting.
See SI text and SI scripts for instruction to include atlases for 3D plotting.

```{r "atlasTab", echo=F, warning=FALSE, message=FALSE}
dd %>%
  mutate(citation = case_when(grepl("dkt", Item) ~ "@dkt")) %>%
  kable(format="latex", escape = TRUE, booktabs = TRUE,
        caption = "Table of currently available atlases in the ggseg and the ggsegExtra R-packages. Polygon and mesh refer to 2D and 3D brain atlase representations.") %>%
  #collapse_rows(1) %>% 
  kable_styling(latex_options = c("scale_down"))
```
